BlueTracer: \\ a Robust API Tracer for Evasive Malware

### 2018-07-09 implementation.tex
Layout Fixed

\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{Figures/BlueTracer.pdf}
\caption{\textit{BlueTracer's high level structure}}
\label{structure}
\end{figure}

@@

In order to bypass this limitation of the Pin's framework, the following strategy was employed:
\begin{enumerate}
\item When initializing the thread local data in \texttt{OnThreadStart}, \texttt{file\_write} is set to point to a function named \texttt{file\_open}.
\item The first time a thread attempts to write data to file, \texttt{file\_open} is invoked.
\item \texttt{file\_open} carries out the following actions:
\begin{enumerate}
\item Opens the tracer's log file (this is safe since the file is not opened in a callback)
\item Sets the obtained file pointer in the thread local data (line 8 of \textit{Listing 3.1})
\item Adds the data to be written in the buffer (which is eventually written to file if the buffer is too small to hold it)
\item Sets \texttt{file\_write} to point to \texttt{buf\_write}, a function which is in charge of just writing data to the buffer and to file.
\end{enumerate}
\item As a result of this, when the thread attempts to write to file again, \texttt{buf\_write} is invoked, thus allowing the thread to just write to file without going through opening the file again.   
\end{enumerate}

@@

In Pin, the Native API is intercepted right before \texttt{sysenter} is executed. This could be inferred by the fact that, in \texttt{TraceSysCallEntry}, the instruction pointer (obtainable via the \texttt{PIN\_GetContextReg} API) contains \texttt{sysenter}'s address. In light of this, in addition to the fact that each Native API calls \texttt{KiFastSystemCall} without setting a stack frame, the application's stack during the execution of \texttt{TraceSysCallEntry} is in the state depicted by \textit{Figure~\ref{sysenter}}.
\\
\begin{figure}[h]
\centering
\includegraphics[width=0.7\textwidth]{Figures/StackMainImage.pdf}
\caption{\textit{Stack before} \texttt{sysenter} \textit{execution}}
\label{sysenter}
\end{figure}     
Having outlined the general situation, it is finally possible to describe how the main image check is carried out:

@@

With this in mind, \texttt{syscallIDs} is built by carrying out the following actions for every element \texttt{iName} in \texttt{AddressOfNames}:

@@

For each Native API, the information related to it is contained in a \texttt{struct} of type \texttt{syscall\_info\_t} (\textit{Listing 3.2}). Most fields of such \texttt{struct} are self-explanatory. In particular, \texttt{num} is a \texttt{struct} storing two values indicating the system call number; these are filled in dynamically by Dr.Memory but are not needed by BlueTracer. Furthermore, the \texttt{flags} field is utilized to notify whether all the details of the Native API are known, as most Native APIs are undocumented. Undoubtedly though, a major role is played by the \texttt{arg} array, which is made up by \texttt{sysinfo\_arg\_t} \texttt{struct}s (\textit{Listing 3.3}) containing the data of the Native API arguments. This array is initialized with size \texttt{MAX\_ARGS\_IN\_ENTRY} (i.e., 18), since a Native API can have at most 18 arguments. 
\\
\begin{lstlisting}[caption={\texttt{struct} containing Native API-related information},captionpos=b]
	typedef struct _syscall_info_t {

@@

Even for \texttt{syscall\_tracer}, each field is self-explanatory. The only one which was not encountered before is the \texttt{syscall\_counter} field. The rationale behind this field is that, in the log, each traced call has a unique integer associated to it, utilized to group together the information relative to the call in post-processing. The next identifier to be assigned is stored in the \texttt{call\_number} field of the thread local storage (\textit{Listing 3.1}). Therefore, \texttt{syscall\_counter} simply contains the unique identifier associated to the Native API in the log. The arguments of a specific Native API are represented via the \texttt{drsys\_arg\_t} struct (\textit{Listing 3.6}).
\texttt{drsys\_arg\_t}'s fields are mostly the same of \texttt{sysinfo\_arg\_t}, with three additions:
\begin{itemize}
\item \texttt{pre}, which is a boolean flag set to \texttt{true} when the logging is occurring before the Native API is executed, i.e., in \texttt{TraceSysCallEntry}, and set to false when the logging is occurring after the Native API is executed, i.e., in \texttt{TraceSysCallExit}.
\item \texttt{enum\_name}, a string containing the symbolic name of the enumeration a named constant belongs to. In most cases, this is the symbolic name associated to the first constant of the enumeration. Such field was introduced to decouple type name and enumeration name.
\item \texttt{value}, an \texttt{ADDRINT}\footnote{The ADDRINT type is defined by Pin and represents a memory address} containing the value of the argument. This is obtained through the \texttt{PIN\_GetSyscallArgument} API, which takes as input the context before the execution of the system call, the system call standard and the ordinal number of the argument whose value is requested.  
\end{itemize}
\begin{lstlisting}[caption={\texttt{struct} representing an argument of the Native API being traced},captionpos=b]
typedef struct _drsys_arg_t {

	// Whether operating pre-call (if true) or post-system call (if false)
	bool pre;

	int ordinal;			    // Parameter Ordinal				
	int size;			        // Size	
	uint flags;			        // SYSARG_ flags
	int type;			        // Type
	const char *type_name;		// Symbolic Name of the arg Type

	// String describing the symbolic name of a named constant's enum
	const char *enum_name;

	// Argument value
	ADDRINT value;

} drsys_arg_t
\end{lstlisting}

@@

\begin{itemize}
\item Then, if the argument's type is primitive (e.g., \texttt{int}, \texttt{bool}, etc.),  \texttt{print\_simple\_value} is invoked to record its value. In particular, the second parameter of this function is a boolean which is set to \texttt{true} if it is wanted to print the argument's value with leading zeros (e.g., if the argument is of pointer type and the address it stores is being logged ) and \texttt{false} otherwise. The way \texttt{print\_simple\_value} operates is quite straightforward:
\begin{enumerate}
\item At the beginning the argument's value is simply logged, with leading zeroes or not depending on the previously mentioned parameter
\item Then, if the argument is a pointer, it is determined if the value it points to needs to be logged. This occurs if the argument is being processed before the Native API execution (i.e., \texttt{pre} is \texttt{true}) and is an input parameter (i.e., \texttt{SYSARG\_READ} is contained in \texttt{flags}) or if the argument is being processed after the Native API execution (i.e., \texttt{pre} is \texttt{false}) and is an output parameter (i.e., \texttt{SYSARG\_WRITE} is contained in \texttt{flags}). The pointed value is obtained through the use of \texttt{PIN\_SafeCopy}, which copies the specified number of bytes from a source memory region to a destination memory region, guaranteeing safe return to the caller even if such regions are inaccessible.  
\end{enumerate}

\vspace{1em}
\begin{lstlisting}[caption={\texttt{print\_arg} function},captionpos=b]
static void print_arg(drsys_arg_t* curr_arg, bluepill_tls* tdata, uint syscall_counter) {

@@

@@

In Microsoft Windows, APIs are implemented as functions provided by a set of dynamically linked libraries. These, also known as DLLs, constitute Microsoft's way of implementing the concept of shared libraries in the Windows operating system. Therefore, the functions they make available can be employed by different applications \cite{DLL}. Windows APIs offer a plethora of functionalities which belong to many different categories, ranging from base services (e.g., file management) to user interface functions and network operations \cite{WinAPI}. In light of this, a Windows API has rich semantic information associated to it and tracing the sequence of APIs a malware calls provides analysts with a high-level view of the sample's behavior. This is why BlueTracer's API tracer component is undoubtedly the most important one. 
\begin{figure}[h]
\centering
\vspace{0.5cm}
\includegraphics[scale=0.662]{Figures/APITracing.pdf}
\caption{\textit{APIs Tracing Workflow}}
\label{API Tracing}
\end{figure}
When tracing APIs, most of the concepts seen in Native API tracing also apply, but there are also significant differences, both in the steps that were carried out and in the Pin APIs which were employed, as it can be observed from the diagram in \textit{Figure~\ref{API Tracing}}.

@@

\textit{drltrace} (already mentioned in Section~\ref{DBI}) comes with its own external configuration file, in which each line contains information relative to an API, as it can be observed from the sample configuration file entry on \textit{Listing 3.9}.      
\vspace{1cm}
\lstset{
    language=C++,
    frame=single, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=none, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}
\begin{lstlisting}[caption={Sample entry in \textit{drltrace} configuration file},captionpos=b]
bool|GetUILanguageInfo|DWORD|wchar*|__out wchar*|__inout DWORD*|__out DWORD*
\end{lstlisting}
\lstset{
    language=C++,
    frame=tb, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=left, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}

@@

In order to extract information relative to the APIs from PyREbox's database, a Python script was employed, which, for every API, instantiates a \texttt{libcall\_info\_t struct} (\textit{Listing 3.10}) in a \texttt{.cpp} file.
\\
\begin{lstlisting}[caption={\texttt{struct} containing API-related information from PyREbox's database},captionpos=b]
typedef struct _libcall_info_t {
	const char* func_name;     			// API name
	int argnum;				   			// Number of Args
	libcall_arg_info_t lib_args[19];	// Array of arguments
} libcall_info_t
\end{lstlisting}

@@

The last operation carried out by \texttt{PrintInfoAfter}, after the API under analysis has been traced, is the re-alignment of the shadow stack, i.e., the deallocation of all the stack entries belonging to APIs which have terminated their execution. In fact, due to the previously mentioned issue involving Pin's instrumentation of routines, it is necessary to manually remove from the shadow stack all the entries relative to the APIs for which the corresponding \texttt{IPOINT\_AFTER} analysis function is not executed. Such shadow stack re-alignment procedure works as follows:

@@

This API allows to register a notification function to execute just before the pinned application changes context due to the reception of an asynchronous event, like for instance a callback or a Windows APC. In BlueTracer, the idea is to use such registered notification function to extract and log information relative to the context change. The notification function in question is of type \texttt{CONTEXT\_CHANGE\_CALLBACK} and the most important arguments it takes as input are:
\begin{itemize}
\item A member of the \texttt{CONTEXT\_CHANGE\_REASON} enumeration indicating the reason for the context change. This is the parameter allowing BlueTracer to distinguish between callbacks and APCs. In fact, in the case of callbacks it is set to \texttt{CONTEXT\_CHANGE\_REASON\_CALLBACK}, while in the case of APCs it is set to \texttt{CONTEXT\_CHANGE\_REASON\_APC}. 
\item The application's register state before the context change. It is \texttt{null} if the context change is caused by a callback.
\item The application's register state after the context change.  
\end{itemize}      
Having outlined the tools made available by Pin to deal with the application's context changes, let us now explain in detail how callbacks tracing and Windows APCs tracing is carried out in BlueTracer.

@@

Windows Asynchronous Procedure Calls (APCs) are employed to alter a thread's ordinary execution path and reroute it to execute some other code. An important concept related to APCs is that, every time an APC is scheduled, it is intended for a specific thread.

APCs are utilized in many situations. Some examples are:

>>>>

\vspace{0.5cm}
\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{Figures/BlueTracer.pdf}
\caption{\textit{BlueTracer's high level structure}}
\label{structure}
\end{figure}

@@

In order to bypass this limitation of the Pin's framework, the following strategy was employed:
\begin{enumerate}
\item When initializing the thread local data in \texttt{OnThreadStart}, \texttt{file\_write} is set to point to a function named \texttt{file\_open}.
\item The first time a thread attempts to write data to file, \texttt{file\_open} is invoked.
\vspace{2cm}
\item \texttt{file\_open} carries out the following actions:
\begin{enumerate}
\item Opens the tracer's log file (this is safe since the file is not opened in a callback)
\item Sets the obtained file pointer in the thread local data (line 8 of \textit{Listing 3.1})
\item Adds the data to be written in the buffer (which is eventually written to file if the buffer is too small to hold it)
\item Sets \texttt{file\_write} to point to \texttt{buf\_write}, a function which is in charge of just writing data to the buffer and to file.
\end{enumerate}
\item As a result of this, when the thread attempts to write to file again, \texttt{buf\_write} is invoked, thus allowing the thread to just write to file without going through opening the file again.   
\end{enumerate}

@@

In Pin, the Native API is intercepted right before \texttt{sysenter} is executed. This could be inferred by the fact that, in \texttt{TraceSysCallEntry}, the instruction pointer (obtainable via the \texttt{PIN\_GetContextReg} API) contains \texttt{sysenter}'s address. In light of this, in addition to the fact that each Native API calls \texttt{KiFastSystemCall} without setting a stack frame, the application's stack during the execution of \texttt{TraceSysCallEntry} is in the state depicted by \textit{Figure~\ref{sysenter}}.
\\
\begin{figure}[h]
\centering
\includegraphics[width=0.7\textwidth]{Figures/StackMainImage.pdf}
\caption{\textit{Stack before} \texttt{sysenter} \textit{execution}}
\label{sysenter}
\end{figure}

Having outlined the general situation, it is finally possible to describe how the main image check is carried out:

@@

%\vspace{0.6cm}
\newpage
With this in mind, \texttt{syscallIDs} is built by carrying out the following actions for every element \texttt{iName} in \texttt{AddressOfNames}:

@@

For each Native API, the information related to it is contained in a \texttt{struct} of type \texttt{syscall\_info\_t} (\textit{Listing 3.2}). Most fields of such \texttt{struct} are self-explanatory. In particular, \texttt{num} is a \texttt{struct} storing two values indicating the system call number; these are filled in dynamically by Dr.Memory but are not needed by BlueTracer. Furthermore, the \texttt{flags} field is utilized to notify whether all the details of the Native API are known, as most Native APIs are undocumented. Undoubtedly though, a major role is played by the \texttt{arg} array, which is made up by \texttt{sysinfo\_arg\_t} \texttt{struct}s (\textit{Listing 3.3}) containing the data of the Native API arguments. This array is initialized with size \texttt{MAX\_ARGS\_IN\_ENTRY} (i.e., 18), since a Native API can have at most 18 arguments. 
\vspace{0.5cm}
\begin{lstlisting}[caption={\texttt{struct} containing Native API-related information},captionpos=b]
	typedef struct _syscall_info_t {

@@

Even for \texttt{syscall\_tracer}, each field is self-explanatory. The only one which was not encountered before is the \texttt{syscall\_counter} field. The rationale behind this field is that, in the log, each traced call has a unique integer associated to it, utilized to group together the information relative to the call in post-processing. The next identifier to be assigned is stored in the \texttt{call\_number} field of the thread local storage (\textit{Listing 3.1}). Therefore, \texttt{syscall\_counter} simply contains the unique identifier associated to the Native API in the log. The arguments of a specific Native API are represented via the \texttt{drsys\_arg\_t} struct (\textit{Listing 3.6}).
\texttt{drsys\_arg\_t}'s fields are mostly the same of \texttt{sysinfo\_arg\_t}, with three additions:
\begin{itemize}
\item \texttt{pre}, which is a boolean flag set to \texttt{true} when the logging is occurring before the Native API is executed, i.e., in \texttt{TraceSysCallEntry}, and set to false when the logging is occurring after the Native API is executed, i.e., in \texttt{TraceSysCallExit}.
\item \texttt{enum\_name}, a string containing the symbolic name of the enumeration a named constant belongs to. In most cases, this is the symbolic name associated to the first constant of the enumeration. Such field was introduced to decouple type name and enumeration name.
\item \texttt{value}, an \texttt{ADDRINT}\footnote{The ADDRINT type is defined by Pin and represents a memory address} containing the value of the argument. This is obtained through the \texttt{PIN\_GetSyscallArgument} API, which takes as input the context before the execution of the system call, the system call standard and the ordinal number of the argument whose value is requested.  
\end{itemize}
\begin{lstlisting}[caption={\texttt{struct} representing an argument of the Native API being traced},captionpos=b]
typedef struct _drsys_arg_t {
	// Whether operating pre-call (if true) or post-system call (if false)
	bool pre;
	int ordinal;			    // Parameter Ordinal				
	int size;			        // Size	
	uint flags;			        // SYSARG_ flags
	int type;			        // Type
	const char *type_name;		// Symbolic Name of the arg Type
	// String describing the symbolic name of a named constant's enum
	const char *enum_name;
	// Argument value
	ADDRINT value;	
} drsys_arg_t
\end{lstlisting}
\newpage

@@

\begin{itemize}
\item Then, if the argument's type is primitive (e.g., \texttt{int}, \texttt{bool}, etc.),  \texttt{print\_simple\_value} is invoked to record its value. In particular, the second parameter of this function is a boolean which is set to \texttt{true} if it is wanted to print the argument's value with leading zeros (e.g., if the argument is of pointer type and the address it stores is being logged ) and \texttt{false} otherwise.

\newpage
\begin{lstlisting}[xleftmargin=\dimexpr-\leftmargini,caption={\texttt{print\_arg} function},captionpos=b]
static void print_arg(drsys_arg_t* curr_arg, bluepill_tls* tdata, uint syscall_counter) {

@@

The way \texttt{print\_simple\_value} operates is quite straightforward:
\begin{enumerate}
\item At the beginning the argument's value is simply logged, with leading zeroes or not depending on the previously mentioned parameter

\item Then, if the argument is a pointer, it is determined if the value it points to needs to be logged. This occurs if the argument is being processed before the Native API execution (i.e., \texttt{pre} is \texttt{true}) and is an input parameter (i.e., \texttt{SYSARG\_READ} is contained in \texttt{flags}) or if the argument is being processed after the Native API execution (i.e., \texttt{pre} is \texttt{false}) and is an output parameter (i.e., \texttt{SYSARG\_WRITE} is contained in \texttt{flags}). The pointed value is obtained through the use of \texttt{PIN\_SafeCopy}, which copies the specified number of bytes from a source memory region to a destination memory region, guaranteeing safe return to the caller even if such regions are inaccessible.  
\end{enumerate}

@@

In Microsoft Windows, APIs are implemented as functions provided by a set of dynamically linked libraries. These, also known as DLLs, constitute Microsoft's way of implementing the concept of shared libraries in the Windows operating system. Therefore, the functions they make available can be employed by different applications \cite{DLL}. Windows APIs offer a plethora of functionalities which belong to many different categories, ranging from base services (e.g., file management) to user interface functions and network operations \cite{WinAPI}. In light of this, a Windows API has rich semantic information associated to it and tracing the sequence of APIs a malware calls provides analysts with a high-level view of the sample's behavior. This is why BlueTracer's API tracer component is undoubtedly the most important one. 
\begin{figure}[h]
\centering
\vspace{1cm}
\includegraphics[scale=0.7]{Figures/APITracing.pdf}
\caption{\textit{APIs Tracing Workflow}}
\label{API Tracing}
\end{figure}
\newpage
When tracing APIs, most of the concepts seen in Native API tracing also apply, but there are also significant differences, both in the steps that were carried out and in the Pin APIs which were employed, as it can be observed from the diagram in \textit{Figure~\ref{API Tracing}}.

@@

\textit{drltrace} (already mentioned in Section~\ref{DBI}) comes with its own external configuration file, in which each line contains information relative to an API, as it can be observed from the sample configuration file entry on \textit{Listing 3.9}.      
\vspace{0.5cm}
\lstset{
    language=C++,
    frame=single, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=none, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}
\begin{lstlisting}[caption={Sample entry in \textit{drltrace} configuration file},captionpos=b]
bool|GetUILanguageInfo|DWORD|wchar*|__out wchar*|__inout DWORD*|__out DWORD*
\end{lstlisting}
\lstset{
    language=C++,
    frame=tb, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=left, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}

@@

\vspace{0.7cm}
\begin{lstlisting}[caption={\texttt{struct} containing API-related information from PyREbox's database},captionpos=b]
typedef struct _libcall_info_t {
	const char* func_name;     			// API name
	int argnum;				   			// Number of Args
	libcall_arg_info_t lib_args[19];	// Array of arguments
} libcall_info_t
\end{lstlisting}

\newpage
In order to extract information relative to the APIs from PyREbox's database, a Python script was employed, which, for every API, instantiates a \texttt{libcall\_info\_t struct} (\textit{Listing 3.10}) in a \texttt{.cpp} file.

@@

The last operation carried out by \texttt{PrintInfoAfter}, after the API under analysis has been traced, is the re-alignment of the shadow stack, i.e., the deallocation of all the stack entries belonging to APIs which have terminated their execution. In fact, due to the previously mentioned issue involving Pin's instrumentation of routines, it is necessary to manually remove from the shadow stack all the entries relative to the APIs for which the corresponding \texttt{IPOINT\_AFTER} analysis function is not executed.
\newpage
Such shadow stack re-alignment procedure works as follows:

@@

This API allows to register a notification function to execute just before the pinned application changes context due to the reception of an asynchronous event, like for instance a callback or a Windows APC. In BlueTracer, the idea is to use such registered notification function to extract and log information relative to the context change. 
\newpage
The notification function in question is of type \texttt{CONTEXT\_CHANGE\_CALLBACK} and the most important arguments it takes as input are:
\begin{itemize}
\item A member of the \texttt{CONTEXT\_CHANGE\_REASON} enumeration indicating the reason for the context change. This is the parameter allowing BlueTracer to distinguish between callbacks and APCs. In fact, in the case of callbacks it is set to \texttt{CONTEXT\_CHANGE\_REASON\_CALLBACK}, while in the case of APCs it is set to \texttt{CONTEXT\_CHANGE\_REASON\_APC}. 
\item The application's register state before the context change. It is \texttt{null} if the context change is caused by a callback.
\item The application's register state after the context change.  
\end{itemize}      
Having outlined the tools made available by Pin to deal with the application's context changes, let us now explain in detail how callbacks tracing and Windows APCs tracing is carried out in BlueTracer.

@@

Windows Asynchronous Procedure Calls (APCs) are employed to alter a thread's ordinary execution path and reroute it to execute some other code. An important concept related to APCs is that, every time an APC is scheduled, it is intended for a specific thread.
\newpage
APCs are utilized in many situations. Some examples are:
